% Related work --- ~0.5 page main text
% Three buckets: (1) CP theory, (2) heteroscedastic UQ, (3) VEP + conformal

\paragraph{Conformal prediction and conditional coverage.}
Split conformal prediction \citep{vovk2005algorithmic,lei2018distribution} provides distribution-free marginal coverage, and Mondrian conformal \citep{vovk2003mondrian} extends this to group-conditional coverage via partitioned calibration; we refer the reader to \citet{angelopoulos2022gentle} for a recent introduction. \citet{barber2020limits} showed that pointwise conditional coverage $P(Y \in \pset(X) \mid X = x)$ is impossible in finite samples without structural assumptions, motivating coarsened conditioning.

\emph{Normalized} nonconformity scores with a learned $\shat$ originate with \citet{papadopoulos2008normalized}, who used kNN-residuals or a separate regressor to scale the absolute residual; \citet{lei2018distribution} §5 supplies a formal analysis of the locally-weighted score $|y - \hat\mu(x)| / \shat(x)$ for regression. \citet{bostrom2020mondrian} systematized Mondrian conformal regression with normalized scores, and the \texttt{crepes} package \citep{bostrom2022crepes} ships an industrial implementation that accepts a difficulty estimator and a Mondrian categorizer as plug-ins. Our score Eq.~\eqref{eq:score} extends this lineage in three respects: (i) it is a binary-classification adaptation of the normalized form (CQR \citep{romano2019cqr} is the regression analog; RAPS \citep{angelopoulos2021raps} is a classification-side adaptive score that does not exploit a learned $\shat$); (ii) the same $\shat$ is embedded inside the Mondrian partition $(y \times \shat\text{-bin})$, not only inside the score; (iii) we provide finite-sample analysis of the bin count $K$ under extreme class imbalance ($\pi_{\min} \approx 0.10$), neither addressed in prior normalized-Mondrian work. \citet{tibshirani2019weighted} established weighted CP under covariate shift, which we invoke to interpret the asymmetric C$\to$M vs M$\to$C transfer in §\ref{sec:experiments}; \citet{gibbs2021adaptive} cover online distribution shift, complementary to our batch chrom-shift bound T4. \citet{barber2023exchangeability} generalized CP beyond exchangeability via weighted residual swaps; we use their Theorem~2 for robust forms T1$'$--T3$'$ and T4.

Three recent lines of work target conditional coverage more directly.
\textbf{RLCP} \citep{hore2025rlcp} uses random localization with kernel weights to achieve feature-space local coverage at rate $O(n^{-2/(d+2)})$; this degrades in high dimensions ($d > 100$), making it impractical for our genomic feature space ($d \approx 7{,}700$).
\citet{dewolf2025conditional} analyze the conditional validity of heteroscedastic Mondrian conformal for regression, establishing when normalized and Mondrian CP adapt correctly to the noise level; our T3--T5 complement their analysis by deriving the optimal Mondrian granularity (bin count $K$) for classification with a \emph{learned} variance head. Concurrently, \citet{zhang2025lwtmcps} introduce LWT-MCPS, a heteroscedasticity-aware Mondrian categorizer that builds bins through a decision tree split by Levene's and Welch's $t$-tests; their setting is regression CRPS over engineering data and the bin partitioner is data-driven, whereas HCCP targets binary classification with $\shat$-quantile bins and provides finite-sample rates that explicitly track the minority-class prevalence $\pi_{\min}$.
\textbf{Self-Calibrating CP} \citep{vanderlaan2024sccp} provides prediction-conditional validity by binning along the calibrated point prediction $\hat\mu$ via Venn--Abers / isotonic calibration, achieving a faster $O(n^{-2/3})$ rate that exploits the monotone structure of isotonic regression \citep{deng2021isotonic} without a Lipschitz score assumption. Their construction is on the orthogonal $\hat\mu$-axis and, as the authors note, does not adapt to mean-independent heteroscedasticity; our T5 conditions on the $\shat$-axis under \ref{ass:ASL} and admits class-imbalance constants ($\pi_{\min}$) that SC-CP, restricted to regression, does not encode. \citet{crps_binning2026} (concurrent) identifies the optimal bin count under covariate-sorted contiguous binning as an open problem in conformal regression; Theorem~\ref{thm:t5oracle} resolves a special case of this open problem for $(y \times \shat\text{-bin})$ Mondrian under class imbalance.

\paragraph{Class-conditional and imbalanced conformal classification.}
\textbf{Class-conditional CP.} \citet{sadinle2019lac} introduced label-conditional CP (the LAC score with separate per-class quantiles), which is the canonical fix for marginal CP's class-imbalance pathology and is functionally equivalent to Mondrian-by-$y$ \citep{vovk2003mondrian}; \citet{shi2024augmentedrank} augment the label rank score for tighter class-wise coverage in many-class settings. \citet{ding2023classcond} cluster classes adaptively to retain class-conditional coverage when per-class calibration sizes are small. \textbf{Imbalanced binary CP.} \citet{xie2025imbalancedcp} (concurrent) study open-set + imbalanced classification with reweighting, providing a $O(n^{-1/2})$ marginal-coverage bound; their analysis does not condition on a heteroscedastic axis and does not let $\pi_{\min}$ enter a local coverage rate.
The empirical observation driving this paper is that none of these methods address the heteroscedastic axis: as our H2H ablation shows (Tab.~\ref{tab:h2h}), \emph{single-axis} Mondrian predictors face a fundamental tradeoff in the imbalanced regime — class-Mondrian preserves $\mathrm{cov}_{|Y=1}$ but loses $\shat$-bin local coverage (gap $0.32$–$0.45$), while $\shat$-Mondrian preserves local coverage but loses $\mathrm{cov}_{|Y=1}$ down to $0.66$ at $\pi_{\min} = 0.10$. HCCP's joint $(y \times \shat\text{-bin})$ Mondrian is, to our knowledge, the first construction simultaneously satisfying both objectives in this regime, with T5 quantifying the price as $O(\sqrt{L_F R / (\pi_{\min} n)})$ — an explicit dependence on $\pi_{\min}$ that prior class-conditional CP and prior imbalanced CP do not provide.

\paragraph{Multi-objective coverage and selective classification.}
Recent work on conformal risk control \citep{angelopoulos2022gentle} and \emph{Pareto testing} \citep{laufer2023paretotest} extends CP to multiple risks; these target a Pareto frontier in the (\emph{risk, set-size}) space rather than a frontier across distinct \emph{coverage} objectives (marginal vs.\ class-conditional vs.\ local). Our analysis is orthogonal: T5 characterises the unique joint partition under which marginal, class-conditional, and $\shat$-bin coverage are all attained simultaneously. We are also \emph{not} a selective-classification framework in the sense of \citet{elyaniv2010selective,geifman2017selective}: HCCP outputs a calibrated prediction set rather than a single label with a learned reject option, and our coverage guarantees apply to every variant rather than to a confident subset.

\paragraph{Heteroscedastic uncertainty quantification.}
Gaussian NLL training for learned variance originates with \citet{nix1994nll}. Deep ensembles \citep{lakshminarayanan2017simple} capture epistemic uncertainty via inter-model disagreement. DEGU \citep{zhou2026degu} pioneered the application of learned heteroscedastic variance heads to genomic UQ, distilling a deep-ensemble teacher into a Gaussian-uncertainty student for MPRA regression. We build directly on this lineage by extending heteroscedastic $\shat$ to (a) variant classification, (b) conformal prediction sets that consume $\shat$ inside the nonconformity score, and (c) finite-sample local coverage with adaptive partition granularity (T5). Our DEGU-lite ablation (§\ref{sec:experiments}) maps out where each $\shat$ source is more informative: ensemble disagreement is comparable to supervised residual regression on Mendelian (monogenic, where epistemic and aleatoric uncertainty are similar in magnitude); supervised $\shat$ captures the aleatoric component that dominates Complex polygenic traits ($4.3\times$ smaller worst-cell gap).

\paragraph{Variant effect prediction and benchmarks.}
TraitGym \citep{benegas2025traitgym} standardizes VEP evaluation on Mendelian and complex trait variants, with a chrom-LOO protocol. Input features include CADD \citep{kircher2014cadd}, Borzoi \citep{linder2025borzoi}, and GPN-MSA \citep{benegas2023gpnmsa}. AlphaGenome \citep{avsec2026alphagenome} is the most recent foundation model for genomics, though it has not been evaluated on TraitGym to date; ProteinGym \citep{notin2023proteingym} is a related large-scale variant fitness benchmark in the protein domain (217 DMS assays, $2.4$M mutants), which we use as a cross-domain stress test for HCCP's coverage machinery (App.~\ref{app:proteingym}). Conformal prediction has been explored in genomic medicine for drug response \citep{papangelou2025cpgenomics} and polygenic risk scores \citep{wang2025cqrprs}, but not for binary variant pathogenicity classification. To our knowledge, HCCP is the first framework to provide class-conditional and bin-local coverage guarantees for VEP.
