% Discussion --- ~0.5 page

\paragraph{Clinical implications.}
HCCP prediction sets are \emph{not} a replacement for expert review; they are a calibrated ``second opinion'' that delineates which variants a VEP model can confidently classify. A singleton set $\pset(x) = \{k\}$ is a confident call; a set containing both labels or neither is an explicit abstention signal. On TraitGym Mendelian, HCCP produces singletons for $\approx$80\% of variants at $\alpha = 0.10$; the remaining 20\% are flagged for expert review, with coverage maintained by construction. This framing complements selective prediction \citep{elyaniv2010selective}: whereas selective prediction maximizes accuracy at a chosen coverage, conformal prediction guarantees coverage at any chosen error rate.

\paragraph{Operating-point selection for polygenic traits.}
The singleton fraction varies sharply with the chosen miscoverage level $\alpha$: $77\%$ at $\alpha = 0.50$, $45\%$ at $\alpha = 0.20$, and $22\%$ at $\alpha = 0.10$ on Complex (App.~\ref{app:alpha}). The low singleton fraction at $\alpha = 0.10$ is not a property of HCCP --- it is a property of the polygenic signal: each individual GWAS variant carries small effect, so the calibrated nonconformity threshold is permissive enough that both classes routinely fit inside the prediction set. Published clinical conformal-prediction deployments operate at higher $\alpha$ where appropriate: \citet{pcos2025conformal} report $41\%$ abstention at the polycystic-ovary-syndrome screening operating point with $\alpha \approx 0.20$, and select that operating point precisely because it balances actionable coverage against expert-review burden. We recommend HCCP users select $\alpha$ by the same calculus: $\alpha = 0.10$ for high-stakes rare-disease classification (where the cost of missing a pathogenic variant dominates), $\alpha = 0.20$ for population-screening triage of complex traits (where the cost of routing every borderline variant to expert review is prohibitive). The coverage guarantee is identical in form across operating points; only the singleton-vs-abstention tradeoff shifts.

\paragraph{Assumptions.}
\ref{ass:A1} (chrom-group exchangeability) is load-bearing. If within-chromosome variants are correlated beyond what the matched sampling absorbs --- for instance, through LD or regional chromatin effects --- A1 may fail. Empirically, the chrom-LOO protocol of TraitGym \citep{benegas2025traitgym} mitigates this by treating each chromosome as an independent fold; our per-chromosome KS audit (App.~\ref{app:data:ks}) shows $10.4\%$ rejection on Complex (close to the $5\%$ null, max KS $0.315$) but $46.2\%$ on Mendelian with the violation concentrated in the highest-$\shat$ positive-class cell where pathogenic variants cluster on a few chromosomes (chr9, chr7, chr3). T3$'$ absorbs this cell-level residual via the TV term, and observed per-chrom coverage gaps remain $\leq 0.073$ as predicted. \ref{ass:ASL} (score-Lipschitz) is required only for the T5 adaptive $K$ derivation; T3 is valid without it. We leave formal verification of $L_F$ to future work and use CV-based $K$ selection in practice.

\paragraph{Cross-dataset transfer scope.}
HCCP's guarantees are conditional on the base predictor retaining discriminative signal and on calibration-fold exchangeability with the test point. C$\to$M transfer preserves both (AUPRC $0.665$, coverage $0.901$, §\ref{sec:experiments}); ProteinGym cross-domain (App.~\ref{app:proteingym}) preserves both with bounded shift. The reverse M$\to$C direction violates both jointly (§\ref{sec:experiments:failure_mode}) and is reported as the boundary case that delineates where HCCP's coverage machinery binds. This boundary is not a method-specific limitation: any conformal procedure inherits the same dependencies on base-predictor competence and calibration-fold exchangeability.

\paragraph{Epistemic versus aleatoric uncertainty.}
The DEGU-lite comparison (§\ref{sec:experiments}) shows that ensemble disagreement ($\shat_{\mathrm{DEGU}}$) and supervised residual regression ($\shat_{\mathrm{HCCP}}$) capture different aspects of uncertainty. On Mendelian (strong signal, well-separated classes), ensemble $\shat$ achieves comparable local coverage. On Complex (weak signal, dominant aleatoric noise), supervised $\shat$ provides $4.3\times$ better worst-cell coverage. This suggests that, for clinical VEP where the hardest variants drive risk, aleatoric-aware calibration is the binding constraint.

\paragraph{Limitations.}
(i) Our cross-dataset TV proxy underestimates the true swap-TV by $\approx$0.036 (honest gap flagged in §\ref{sec:experiments}); closing this requires access to an oracle model.
(ii) The T5 rate $O(n^{-1/2})$ assumes equiprobable bins; non-uniform partitions could improve constants but not the rate.
(iii) TraitGym's matched sampling design controls for consequence type, MAF, and TSS distance, which suppresses distributional heterogeneity that would be present in a clinical cohort.
(iv) The oracle bin count $K^\star$ derived in Theorem~\ref{thm:t5oracle} is a \emph{large-sample upper bound}; in the finite-sample regime ($n \leq 11{,}400$ on TraitGym) the binding constraint is the per-cell floor $n_{k,b} \geq \lceil 1/\alpha \rceil$, and CV-selected $\hat K_{\mathrm{CV}}$ is substantially smaller than $K^\star$ (App.~\ref{app:tables:ksweep}). We recommend using $\hat K_{\mathrm{CV}}$ in practice and treating $K^\star$ as asymptotic guidance.
(v) We compare against a GBM-backbone reimplementation of DEGU (DEGU-lite, §\ref{sec:experiments}) to isolate the conformal contribution from the backbone choice. A full TF/Keras CNN end-to-end port of DEGU \citep{zhou2026degu} to TraitGym is out of scope: DEGU's CNN consumes 200\,bp DNA windows designed for MPRA regression, whereas variant pathogenicity depends on gene annotation, long-range regulatory context, and phylogenetic signal that fit in pre-computed features (CADD, Borzoi 524\,kb context, GPN-MSA) but not in 200\,bp inputs.

\paragraph{Broader impact.}
\textbf{Clinical deployment risks.} VEP systems inform consequential decisions (cascade testing, reproductive choices, preventive surgery). Miscalibrated confidence can silently push users toward overconfident rare-disease diagnoses or dismissive classification of ambiguous variants. HCCP provides \emph{calibrated} abstention, but abstention does not \emph{guarantee} clinical safety: the remaining $\approx 20\%$ singletons on Mendelian still carry a $10\%$ finite-sample miscoverage floor, and the $\{0, 1\}$ abstention signal must be paired with a human expert review policy to translate coverage into clinical utility.

\textbf{Population representativeness.} TraitGym is built from ClinVar (Mendelian) and the GWAS catalog (Complex), both of which over-represent individuals of European ancestry. Coverage guarantees derived on TraitGym do not automatically transfer to under-represented populations. Operationalising HCCP in a diverse clinical setting requires re-calibration on population-stratified folds, which is a direct application of the Mondrian machinery (partition by ancestry group). A fourth Mondrian axis $(y \times \shat\text{-bin} \times \text{ancestry-bin})$ would extend T3 to ancestry-conditional coverage at the cost of an $|\mathcal{A}|$-fold reduction in per-cell calibration size; for $|\mathcal{A}| = 5$ super-populations with $\pi_{\min} = 0.10$, Theorem~\ref{thm:t5oracle} implies $K^\star \approx \sqrt{L_F R \pi_{\min} n / |\mathcal{A}|}$, so the $\shat$-bin count shrinks by $\sqrt{|\mathcal{A}|}$ to maintain per-cell sample size. Empirical validation pending ancestry-labelled VEP benchmarks similar to UK Biobank stratified PRS evaluation \citep{wang2025cqrprs}; we flag this as the most operationally relevant gap between TraitGym-derived guarantees and clinical deployment.

\textbf{Downstream misuse.} A singleton prediction $\pset(x) = \{1\}$ is a \emph{necessary but not sufficient} condition for pathogenicity: the ACMG/AMP consensus \citep{richards2015acmg} explicitly treats computational predictions as evidence category PP3/BP4 (supporting), not as stand-alone classification. Positioning HCCP singletons as diagnostic is incorrect use; positioning them as one calibrated evidence line in a multi-evidence framework is the intended use. We encourage downstream integrators to document which prediction sets feed which ACMG evidence tier and to track population-stratified coverage post-deployment.
