% Appendix B: Data, features, and training details

\subsection{TraitGym benchmark}
\label{app:data:traitgym}

We use TraitGym \citep{benegas2025traitgym}, a standardized evaluation suite for variant effect prediction. Two matched datasets:
\begin{itemize}
    \item \textbf{Mendelian (matched\_9)}: 338 ClinVar pathogenic variants + 3,042 matched benign controls = 3,380 total across 113 OMIM phenotypes. Matching on consequence type, minor allele frequency (MAF), and transcription start site (TSS) distance (9 strata).
    \item \textbf{Complex (matched\_9)}: 1,140 GWAS-significant variants + 10,260 matched controls = 11,400 total across 29 complex traits.
\end{itemize}

\paragraph{Chrom-LOO protocol.}
Test chromosomes: $\{17, 18, 19, 20, 21, 22, X\}$ (TraitGym default). For each test chrom $c^\star$, the remaining chromosomes form the train/calibration fold. Primary metric: $\mathrm{AUPRC}_{\mathrm{chrom\text{-}weighted}} = \sum_c w_c \cdot \mathrm{AUPRC}_c$ with $w_c = n_c / \sum_{c'} n_{c'}$.

\subsection{Input features}
\label{app:data:features}

\begin{table}[h]
\centering
\small
\caption{Features used in HCCP aggregator. All pre-computed as per TraitGym conventions.}
\begin{tabular}{llr}
\toprule
Feature & Source & Dim \\
\midrule
CADD v1.7 & \citet{kircher2014cadd} & 1 \\
Borzoi\_L2 & \citet{linder2025borzoi} zero-shot L2-norm & 3{,}846 \\
Borzoi\_L2\_L2 & \citet{linder2025borzoi} L2-of-L2 & 3{,}846 \\
GPN-MSA\_LLR & \citet{benegas2023gpnmsa} log-likelihood ratio & 1 \\
GPN-MSA\_absLLR & \citet{benegas2023gpnmsa} absolute LLR & 1 \\
GPN-MSA\_InnerProducts & \citet{benegas2023gpnmsa} inner products & 36 \\
AlphaGenome & \citet{avsec2026alphagenome} RNA\_SEQ $+$ DNASE $\times$ 8 tissues & 140 \\
\midrule
\textbf{CADD+GPN-MSA+Borzoi (full)} & & \textbf{7{,}731} \\
\textbf{CADD+GPN-MSA+Borzoi+AlphaGenome} & & \textbf{7{,}871} \\
\textbf{CADD+Borzoi} & & \textbf{7{,}693} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Aggregator hyperparameters}
\label{app:data:hparams}

\paragraph{Base classifier $\phat$.}
\texttt{HistGradientBoostingClassifier} (scikit-learn 1.4): \texttt{max\_depth=2}, \texttt{max\_iter=100}, \texttt{learning\_rate=0.1}, \texttt{class\_weight=balanced}, \texttt{random\_state=42}, \texttt{early\_stopping=False}. Feature imputation: mean via \texttt{SimpleImputer}. No feature scaling (GBM is scale-invariant).

\paragraph{AlphaGenome features.}
We extract predictions from the AlphaGenome API \citep{avsec2026alphagenome} on $16$\,kb windows centered on each variant, across 8 representative tissue ontology terms (brain, liver, heart, lung, kidney, colon, bone marrow, pancreas) for \texttt{RNA\_SEQ} and \texttt{DNASE} output tracks. For each $(\text{output}, \text{tissue})$ pair we record (a) window-mean $\log(1 + x)$ of ref and alt predictions, (b) center-position alt$-$ref delta in log space, and (c) maximum $|$alt$-$ref$|$ across the window, yielding $2 \cdot 8 \cdot (2 + 2) = 64$ entries per output type, 128 combined --- plus 12 aggregated cross-tissue summaries for 140 total. Extraction used the official Python client with 8 parallel workers; 97\% (Mendelian) / 98\% (Complex) of variants returned complete feature vectors in $\approx 8$--$19$ minutes total wall time. Remaining variants (near chromosome boundaries or in inaccessible regions) receive NaN features, which \texttt{HistGradientBoostingClassifier} handles natively.

\paragraph{Heteroscedastic head $\shat$.}
\texttt{HistGradientBoostingRegressor} with identical backbone parameters. Target: $r_i = |Y_i - \phat^{(-C_i)}(X_i)|$ (chrom-LOO OOF residuals). Loss: squared error (Gaussian NLL approximation). Floor: $\shat(x) = \max(\shat_{\mathrm{raw}}(x), \epsilon)$ with $\epsilon = 10^{-4}$.

\paragraph{Conformal calibration.}
Score: $s(x, y) = |y - \phat(x)| / (\shat(x) + \epsilon)$. Bin count: $K = \hat K(c_{\mathrm{outer}})$ via nested chrom-LOO (§\ref{sec:theory:t5}); fallback $K = 5$ for fixed-K baselines (Tab.~\ref{tab:main}). Cell minimum $n_{k,b} \geq 5$; fallback to class-pooled $\qhat_k$ otherwise. Miscoverage: $\alpha = 0.10$.

\paragraph{DEGU-lite ensemble.}
$M = 10$ GBM members, each with bootstrap resampling (resample with replacement) and unique \texttt{random\_state}. Ensemble uncertainty: $\shat_{\mathrm{DEGU}}(x) = \mathrm{std}_{m=1}^M \phat_m(x)$. Same backbone as aggregator.

\subsection{Within-cell KS audit}
\label{app:data:ks}

To assess A2-cell (score stationarity within Mondrian cells across chromosomes), for each $(k, b)$ cell we compared each chromosome's score distribution against the pooled distribution of all other chromosomes in the same cell via two-sample Kolmogorov--Smirnov tests. Cells with $n \geq 20$ per-chrom were retained (52 tests on Mendelian, 125 on Complex, matching the $\approx 155$-test budget reported in earlier drafts aggregated across datasets).

\begin{table}[h]
\centering
\small
\caption{Per-chromosome A2-cell KS audit (CADD+GPN-MSA+Borzoi). A perfectly stationary score law would reject at the nominal $5\%$ rate. The score law $s(x,y) = |y - \phat(x)| / (\shat(x) + \epsilon)$ is used.}
\label{tab:ks_audit}
\begin{tabular}{lrrrrr}
\toprule
Dataset & $n_{\text{tests}}$ & Reject @ $p{<}0.05$ & KS max & KS mean & KS median \\
\midrule
Mendelian & 52  & 46.2\% (24/52)  & 0.434 & 0.207 & 0.163 \\
Complex   & 125 & 10.4\% (13/125) & 0.315 & 0.113 & 0.102 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Interpretation.} A2-cell is approximately satisfied on Complex (10.4\% vs 5\% null, max KS $0.315$) but noticeably violated on Mendelian (46.2\%, max KS $0.434$). The violation concentrates in the highest-$\shat$ positive-class cell $(y{=}1, b{=}4)$ --- chromosomes chr9, chr7, chr3, chr19, and chrX each show KS $\geq 0.40$ at this cell, consistent with chromosome-specific pathogenic variant concentration. The robust form T3$'$ absorbs this violation via the chrom-TV term: our observed per-chrom max marginal-coverage gap of $0.100$ (chr6, Tab.~\ref{tab:perchrom}) is within the $\bar\delta_{\mathrm{TV}}^{(k,b)}$ bound consistent with max KS $\leq 0.44 \Rightarrow d_{\mathrm{TV}} \leq 0.22$. Full per-chromosome breakdowns: \texttt{outputs/ks\_audit/*\_ks\_per\_chrom\_n20.json}.

\paragraph{$K$-invariance of the rejection rate.} To confirm that the 46.2\% rejection on Mendelian is a dataset-intrinsic chrom-TV signal rather than a binning artefact at the legacy default $K = 5$, we re-run the per-chromosome KS audit on Mendelian at $K_{\mathrm{eval}} \in \{2, 3, 5, 8, 10\}$ with the same min-per-chrom-cell threshold of $20$ (Tab.~\ref{tab:ks_kinvariance}). The rejection rate stays in the $40$--$46\%$ band across all five $K$ values, and the maximum cell-by-chrom KS statistic stays in $[0.434, 0.565]$. T3$'$ is therefore the binding operational guarantee on Mendelian regardless of partition granularity. Source: \texttt{outputs/ks\_audit\_kcv/mendelian\_K\{2,3,5,8,10\}\_n20.json}.

\begin{table}[h]
\centering
\small
\caption{Mendelian per-chromosome A2-cell KS audit $K$-sweep at min-per-chrom-cell$=20$. Rejection rate is the fraction of (cell, chromosome) pairs at which $p < 0.05$.}
\label{tab:ks_kinvariance}
\begin{tabular}{lrrr}
\toprule
$K_{\mathrm{eval}}$ & $n_{\text{tests}}$ & Reject @ $p{<}0.05$ & KS max \\
\midrule
2 & 37 & 45.9\% (17/37) & 0.565 \\
3 & 45 & 40.0\% (18/45) & 0.524 \\
5 & 52 & \textbf{46.2\% (24/52)} & 0.434 \\
8 & 58 & 43.1\% (25/58) & 0.492 \\
10 & 60 & 45.0\% (27/60) & 0.558 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Leakage sanity audit}
\label{app:data:leakage}

Because \texttt{HistGradientBoostingClassifier} at \texttt{max\_depth=2} with \texttt{early\_stopping=False}, default \texttt{subsample=1.0}, and \texttt{max\_features=None} contains no active random source, the classifier output is \emph{deterministic given the training data} and \texttt{random\_state} produces no variance. We view this as an algorithmic property (sklearn 1.4 documentation confirms that \texttt{random\_state} controls only early-stopping validation splits, row subsampling, and binning tie-breaking, none of which activate under our configuration), but algorithmic determinism is observationally indistinguishable from \emph{data leakage} absent additional tests. We therefore report three orthogonal sanity checks on CADD+GPN-MSA+Borzoi Mendelian and CADD+Borzoi Complex.

\begin{table}[h]
\centering
\small
\caption{Leakage sanity audit. Test A trains on shuffled labels and evaluates on the original labels: under leakage, shuffling would leave AUPRC high; we observe collapse toward the positive-class prior ($\approx 0.10$). Test B forces row-level stochasticity via 80\% random row masking with seeds $\{42,7,2024\}$ and confirms \texttt{random\_state} \emph{does} produce variance under active randomness --- but less than $0.003$ AUPRC std, reflecting TraitGym's strong feature signal rather than a bug. Test C audits the chrom-LOO split.}
\label{tab:leakage_sanity}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{llrr}
\toprule
Test & Condition & Mendelian AUPRC & Complex AUPRC \\
\midrule
Reference          & seed 42, no perturbation                  & 0.900          & 0.350          \\
\midrule
A: label shuffle   & 10\% labels flipped                        & 0.898          & 0.327          \\
A: label shuffle   & 30\% labels flipped                        & 0.776          & 0.257          \\
A: label shuffle   & \textbf{50\% labels flipped}               & \textbf{0.147} & \textbf{0.126} \\
\midrule
B: subsample+seed  & row-mask 0.8 (std over 3 seeds)            & 0.0006 (std)   & 0.0027 (std)   \\
\midrule
C: chrom-LOO audit & train $\cap$ test, folds 19--22            & 0 overlap      & 0 overlap      \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Interpretation.}
(i) \emph{Test A is the critical test}: on both datasets, training with 50\% random label corruption collapses test AUPRC from 0.90/0.35 to 0.147/0.126, approaching the positive-class base rate 0.10. This is inconsistent with data leakage --- a leaked model would retain its AUPRC regardless of training labels --- and consistent with the model genuinely learning label signal.
(ii) Test B shows that when stochasticity is forcibly introduced via row subsampling, seed variance is strictly positive ($0.0006$ on Mendelian, $0.0027$ on Complex). The asymmetry is biologically interpretable: Complex has weaker signal (AUPRC $0.35$ vs $0.90$), so the model is more sensitive to training perturbations, as expected. The fact that both stds are small ($< 0.003$) reflects TraitGym's strong pre-computed features (CADD, Borzoi, GPN-MSA) --- the majority of predictive signal comes from the features themselves, not the boosting trajectory.
(iii) Test C confirms the chrom-LOO split is structurally clean.
Together, the three tests establish that the zero seed-variance observation in the main text is an algorithmic property, not a leakage artifact, and that chromosome-level bootstrap (App.~\ref{app:data:compute}) is the appropriate uncertainty quantifier.

\subsection{Compute}
\label{app:data:compute}

All experiments on a single machine with 4$\times$ NVIDIA T4 GPUs (SM 7.5, no BF16). The HCCP pipeline (aggregator + $\shat$ head + conformal calibration) runs entirely on CPU via scikit-learn.

\begin{table}[h]
\centering
\small
\caption{Wall-clock cost of the full HCCP pipeline and main baselines, single chrom-LOO sweep on TraitGym. CPU times on 16-core Xeon; GPU on a single T4. Memory peak excludes Python overhead. $\phat$ training is per feature set; $\shat$ training is on chrom-LOO OOF residuals; Mondrian calibration is reported per $K$. DEGU-lite shares the $\phat$ backbone. The $B = 200$ chromosome-level bootstrap (Tab.~\ref{tab:main}) reuses cached calibration scores and adds $\sim 30$ CPU-min total.}
\label{tab:compute}
\setlength{\tabcolsep}{5pt}
\begin{tabular}{llrrr}
\toprule
Stage & Backend & Mendelian & Complex & Mem peak \\
\midrule
$\phat$ training (HistGB)                    & CPU 16-core       & $\approx 20$ min & $\approx 60$ min & 4 GB \\
$\shat$ training (HistGB Gaussian-NLL)        & CPU 16-core       & $\approx 15$ min & $\approx 45$ min & 4 GB \\
Mondrian conformal calibration                & CPU single-thread & $< 5$ min        & $< 5$ min        & 1 GB \\
Total HCCP (single feature set, all chroms)   & CPU 16-core       & $< 4$ CPU-h      & $< 4$ CPU-h      & 4 GB \\
\midrule
DEGU-lite ($M = 10$ bootstrap GBM)            & CPU 16-core       & $\approx 35$ min & $\approx 75$ min & 6 GB \\
\midrule
$B = 200$ chrom-bootstrap                     & CPU single-thread & $\approx 12$ min & $\approx 18$ min & 1 GB \\
Multi-seed sweep ($\{42, 7, 2024\}$)          & CPU 16-core       & $\approx 12$ min & $\approx 18$ min & 4 GB \\
\bottomrule
\end{tabular}
\end{table}

Figure and proof generation are negligible (matplotlib + LaTeX).
