% Appendix C: Detailed per-partition results

\subsection{Per-chromosome coverage}
\label{app:tables:perchrom}

Table~\ref{tab:perchrom} reports per-chromosome coverage for the HCCP Mondrian predictor (CADD+GPN-MSA+Borzoi, $\alpha = 0.10$, $K = 5$).

\begin{table}[h]
\centering
\small
\caption{Per-chromosome coverage under Mondrian-$(y \times \shat\text{-bin})$, Mendelian dataset. Target: 0.90. Chromosomes 17--22, X are test chroms under chrom-LOO.}
\label{tab:perchrom}
\begin{tabular}{lrrr}
\toprule
Chrom & $n$ & Coverage & $|\mathrm{gap}|$ \\
\midrule
1  & 210 & 0.857 & 0.043 \\
2  & 230 & 0.896 & 0.004 \\
3  & 155 & 0.945 & 0.045 \\
5  & 100 & 0.850 & 0.050 \\
6  & 150 & 0.800 & 0.100 \\
7  & 210 & 0.943 & 0.043 \\
8  & 140 & 0.943 & 0.043 \\
9  & 240 & 0.842 & 0.058 \\
10 & 190 & 0.879 & 0.021 \\
11 & 160 & 0.944 & 0.044 \\
12 & 150 & 0.933 & 0.033 \\
13 & 210 & 0.848 & 0.052 \\
14 & 100 & 0.900 & 0.000 \\
16 & 80  & 0.913 & 0.013 \\
17 & 150 & 0.900 & 0.000 \\
19 & 200 & 0.828 & 0.073 \\
20 & 50  & 0.900 & 0.000 \\
22 & 20  & 0.950 & 0.050 \\
X  & 250 & 0.958 & 0.058 \\
\midrule
\textbf{Overall} & \textbf{3380} & \textbf{0.902} & \textbf{0.002} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{K-sweep ablation}
\label{app:tables:ksweep}

Table~\ref{tab:ksweep} validates Theorem~\ref{thm:t5oracle} (T5.1) empirically by sweeping $K \in \{2, 3, 5, 8, 10, 15, 20, 30\}$. For each dataset the plug-in Lipschitz estimate is $\hat L_F \approx 49.4$ (Mendelian) and $21.8$ (Complex) via 100-batch finite differences on $\shat(X)$; the resulting oracle bin count $K^\star = \lfloor\sqrt{L_F R \pi_{\min} n}\rfloor$ is $88$ (Mendelian) and $110$ (Complex), neither of which is realizable because the finite-sample variance term $K/(\pi_{\min} n)$ dominates long before the stationarity bias term $L_F R / K$ crosses it; the cross-validated $\hat K_{\mathrm{CV}}$ picks the binding constraint instead, yielding $3$ (Mendelian) and $2$ (Complex). The fallback column reports the number of $(k,b)$ cells that triggered pooled class-conditional backoff ($n_{k,b} < 5$, §\ref{sec:method}).

\begin{table}[h]
\centering
\small
\caption{$K$-sweep on Mendelian and Complex (CADD+GPN-MSA+Borzoi, $\alpha = 0.10$, single seed). Worst-cell gap = $\max_{k,b} |\mathrm{cov}_{k,b} - 0.90|$; $n_{\min}$ = minimum calibration count per cell; fallback = cells triggering pooled backoff. Bolded rows are $\hat K_{\mathrm{CV}}$.}
\label{tab:ksweep}
\begin{tabular}{rrrrrrrrr}
\toprule
& \multicolumn{4}{c}{Mendelian} & \multicolumn{4}{c}{Complex} \\
\cmidrule(lr){2-5} \cmidrule(lr){6-9}
$K$ & Worst gap & Mean gap & $n_{\min}$ & fallback & Worst gap & Mean gap & $n_{\min}$ & fallback \\
\midrule
2  & 0.233 & 0.059 & 6  & 0  & \textbf{0.006} & \textbf{0.002} & 237 & 0 \\
\textbf{3}  & \textbf{0.047} & \textbf{0.012} & 19 & 1  & 0.010 & 0.004 & 131 & 0 \\
5  & 0.285 & 0.041 & 13 & 2  & 0.042 & 0.007 & 52  & 0 \\
8  & 0.627 & 0.067 & 11 & 5  & 0.058 & 0.010 & 19  & 0 \\
10 & 0.700 & 0.059 & 10 & 6  & 0.208 & 0.017 & 13  & 0 \\
15 & 0.733 & 0.049 & 6  & 9  & 0.600 & 0.034 & 10  & 0 \\
20 & 0.733 & 0.047 & 6  & 14 & 0.600 & 0.030 & 10  & 1 \\
30 & 0.344 & 0.031 & 5  & 22 & 0.567 & 0.034 & 9   & 2 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Reading the table.} (i) On Mendelian, the worst-cell gap falls 5$\times$ from $K{=}2$ to $K{=}3$ (0.233 $\to$ 0.047), then rises monotonically until $K{=}15$ as fallback triggers accumulate. This is the bias--variance tradeoff of Eq.~\ref{eq:G_K} realized empirically: at $K{=}3$ the stationarity bias is still dominant; beyond $K{=}3$ the finite-sample variance $K/(\pi_{\min} n)$ takes over, amplified by cells dropping below $n_{k,b} \geq \lceil 1/\alpha \rceil = 10$. (ii) On Complex, the minority class has $n_1 = 1{,}140$ vs Mendelian's $338$, so variance dominance is pushed out to $K \geq 10$; the sweet spot is $K \in \{2, 3\}$. (iii) The number of fallback cells is a mechanical audit of the theoretical assumption that $n_{k,b} \geq \lceil 1/\alpha \rceil$; on Mendelian it hits $22/60$ at $K{=}30$, confirming the sample-size floor as the binding constraint on $K$ in this regime.

\paragraph{$L_F$ estimator sensitivity.}
The oracle bin count $K^\star = \lfloor\sqrt{L_F R \pi_{\min} n}\rfloor$ depends on the plug-in Lipschitz estimate $\hat L_F$. We compare four estimators in Table~\ref{tab:LF_estimators}: the \emph{legacy max-of-ratios} over adjacent pilot bins (used in earlier drafts), a \emph{median-of-ratios} variant, a \emph{p90-of-ratios} variant, and an \emph{empirical inversion} that solves $G(K) = \hat L_F R/K + K/(\pi_{\min} n)$ for $\hat L_F$ at each observed $K$ and takes the median. The estimators span $[5.8, 49.7]$ on Mendelian and $[2.4, 21.8]$ on Complex. All four yield a theoretical $K^\star$ substantially above the cross-validated $\hat K_{\mathrm{CV}} \in \{2, 3\}$; the gap is driven by a finite-sample effect the T5 bound does not model: when $n_{k,b} < \lceil 1/\alpha\rceil$, the implementation falls back to the pooled class-conditional quantile $\hat q_k$ (T2), which is uniform across $\shat$-bins and therefore \emph{widens} the worst-cell gap non-monotonically in $K$. We interpret $\hat K_{\mathrm{CV}}$ as the CV-calibrated lower bound and $K^\star$ as the asymptotic upper bound; practitioners should choose $\hat K_{\mathrm{CV}}$ in the finite-sample regime and expect the two to converge as $n \to \infty$.

\begin{table}[h]
\centering
\small
\caption{Four $L_F$ estimators and their oracle $K^\star = \lfloor\sqrt{L_F R \pi_{\min} n}\rfloor$ and $G(K^\star) \leq 2\sqrt{L_F R / (\pi_{\min} n)}$ bounds. The cross-validated $\hat K_{\mathrm{CV}}$ is reported for comparison.}
\label{tab:LF_estimators}
\begin{tabular}{llrrr}
\toprule
Estimator & Dataset & $\hat L_F$ & $K^\star$ & $G(K^\star)$ bound \\
\midrule
legacy max-of-ratios   & Mendelian & 49.7 & 88  & 0.525 \\
median-of-ratios       & Mendelian & 5.8  & 30  & 0.179 \\
p90-of-ratios          & Mendelian & 23.5 & 61  & 0.362 \\
empirical inversion    & Mendelian & 12.3 & 44  & 0.261 \\
\midrule
\multicolumn{2}{l}{\emph{cross-validated} $\hat K_{\mathrm{CV}}$ (Mendelian)} &   & \textbf{3} & 0.047 (observed) \\
\midrule
legacy max-of-ratios   & Complex   & 21.8 & 110 & 0.194 \\
median-of-ratios       & Complex   & 6.9  & 62  & 0.109 \\
p90-of-ratios          & Complex   & 19.5 & 104 & 0.183 \\
empirical inversion    & Complex   & 2.4  & 36  & 0.065 \\
\midrule
\multicolumn{2}{l}{\emph{cross-validated} $\hat K_{\mathrm{CV}}$ (Complex)}   &   & \textbf{2} & 0.006 (observed) \\
\bottomrule
\end{tabular}
\end{table}

\subsection{DEGU-lite vs HCCP head-to-head}
\label{app:tables:degu}

Table~\ref{tab:degu} compares HCCP's supervised $\shat$ against DEGU-lite's ensemble $\shat$ under identical Mondrian-$(y \times \shat\text{-bin})$ conformal calibration and feature sets.

\begin{table}[h]
\centering
\small
\caption{HCCP vs DEGU-lite: same features (CADD+GPN-MSA+Borzoi), same Mondrian partition, different $\shat$ source. Worst gap = max per-$\shat$-bin deviation from target 0.90.}
\label{tab:degu}
\begin{tabular}{llrrrr}
\toprule
$\shat$ source & Dataset & Cov & Cov$|_{Y{=}1}$ & Worst gap & Mean gap \\
\midrule
HCCP (learned)    & Mendelian & 0.902 & 0.905 & 0.041 & 0.022 \\
DEGU (ensemble)   & Mendelian & 0.901 & 0.917 & \textbf{0.033} & \textbf{0.020} \\
\midrule
HCCP (learned)    & Complex   & 0.901 & 0.908 & \textbf{0.013} & \textbf{0.007} \\
DEGU (ensemble)   & Complex   & 0.901 & 0.907 & 0.057 & 0.028 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Interpretation.}
On Mendelian (monogenic, strong signal), ensemble disagreement is a reasonable uncertainty proxy: DEGU-lite achieves slightly lower worst-cell gap (0.033 vs.\ 0.041). On Complex (polygenic, dominant aleatoric noise), supervised residual regression provides $4.3\times$ better worst-cell coverage (0.013 vs.\ 0.057). This confirms the epistemic--aleatoric decomposition discussed in §\ref{sec:discussion}: where model disagreement and data noise contribute equally, both $\shat$ sources suffice; where data noise dominates (Complex traits), the supervised head captures what ensembles cannot.

\subsection{Cross-dataset transfer}
\label{app:tables:crossdataset}

\begin{table}[h]
\centering
\small
\caption{Cross-dataset transfer (CADD+Borzoi). M$\to$C: train on Mendelian, test on Complex. C$\to$M: reverse. T4 proxy bound = $1 - \alpha - d_{\mathrm{TV}}$.}
\label{tab:crossdataset}
\begin{tabular}{lrrrrr}
\toprule
Direction & AUPRC & Coverage & Gap & TV proxy & T4 bound \\
\midrule
M$\to$C & 0.187 & 0.738 & $-0.162$ & 0.158 & 0.742 \\
C$\to$M & 0.654 & 0.901 & $+0.001$ & 0.158 & 0.742 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Directional asymmetry.}
C$\to$M transfer preserves coverage (0.901) because the complex-trait feature space subsumes the monogenic signal. M$\to$C collapses (0.738) because monogenic variants occupy a narrow subspace. The Barber T4 proxy bound (0.742) is tight for M$\to$C (observed 0.738, within 0.004) and conservative for C$\to$M.
