% Appendix G: Open Targets cross-platform replication

To rule out the possibility that HCCP's $\shat$-bin gap advantage is an
artifact of TraitGym's pre-computed pathogenicity feature pipeline (CADD +
GPN-MSA + Borzoi), we replicate the headline $H_2H$ on a third dataset
constructed from \textbf{Open Targets Platform 25.06} \citep{ochoa2023opentargets}
fine-mapped GWAS credible sets, using a completely independent feature
pipeline (gnomAD allele frequency + VEP consequence + AlphaMissense + SIFT +
GERP + LOFTEE).

\paragraph{Dataset construction.}
From the 2.6M credible sets in OT 25.06, we keep GWAS studies with credible
set size $\in [5, 200]$ and binarize variants by posterior inclusion
probability (PIP): variants with PIP $\geq 0.5$ are positives ($24{,}895$
unique high-PIP variants survive the AF-non-NaN filter). For each positive,
we sample $9$ AF-matched random hg38 variants from the gnomAD pool
\emph{outside any OT credible set}, with allele frequency within
$\pm 0.05$ of the positive (TraitGym matched-9 protocol). After
deduplication on \texttt{variantId}, we subsample to
$1{,}140$ positives + $10{,}260$ controls
to mirror TraitGym Complex matched\_9 ($n = 11{,}400$, $\pi_{+} = 0.10000$),
with chrom-LOO test split chr $\{17, 18, 19, 20, 21, 22, X\}$ matching
\S\ref{sec:experiments}. Variant overlap with TraitGym Complex / Mendelian
test sets is $\approx 0$ at $(\mathrm{chrom}, \mathrm{pos}, \mathrm{ref}, \mathrm{alt})$
resolution and $1{,}846 / 25{,}068$ at $(\mathrm{chrom}, \mathrm{pos})$
resolution; OT positives are predominantly disjoint from TraitGym.

\paragraph{Feature pipeline.}
We use $13$ leak-free features extracted from OT \texttt{variant.parquet}:
gnomAD allele frequencies (\texttt{af\_overall}, \texttt{af\_nfe}, \texttt{af\_max}),
GERP conservation, AlphaMissense, SIFT, FoldX, LOFTEE LoF flag, VEP-derived
LoF flag, ordinal consequence severity (0--15 per the SO hierarchy), and
three feature-presence flags (NaN-handling). \texttt{pip} and other
fine-mapping outputs are excluded as label-leaking. The aggregator is
HistGradientBoosting (\texttt{max\_depth=3}, \texttt{max\_iter=300},
\texttt{class\_weight="balanced"}) trained chrom-LOO with OOF residuals for
the heteroscedastic head, identical to the TraitGym recipe. Full pool OOF
$\mathrm{AUPRC} = 0.397$, $\mathrm{AUROC} = 0.831$ — a hard regime
comparable to TraitGym Complex ($\mathrm{AUPRC} = 0.353$).

\paragraph{Headline H2H result.}
Tab.~\ref{tab:open_targets_h2h} reports the four-method comparison at the
recommended operating point ($K_{\mathrm{eval}} = 5$, $\alpha = 0.10$) under
the same chrom-LOO + nested-CV $\hat K(c_{\mathrm{outer}})$ protocol used
for TraitGym (\S\ref{sec:experiments:h2h}, Tab.~\ref{tab:cp_baselines}).

\begin{table}[h]
\centering
\small
\caption{Open Targets matched\_9 H2H, $K_{\mathrm{eval}} = 5$, $\alpha = 0.10$.
HCCP partition $K_{\mathrm{HCCP}}$ selected per outer fold by nested
chrom-LOO. Paired bootstrap on $B = 200$ chrom-resamples (seed 42); win\,\%
counts resamples on which HCCP attained a strictly smaller $\shat$-bin gap
than the baseline. The three baselines all hit $\shat$-bin gap $0.900$, the
worst-cell maximum-deviation ceiling (one minority cell at near-zero
coverage); the $16.7\times$ ratio is therefore better read as ``HCCP holds
the gap floor while the baselines saturate the ceiling'' than as a
multiplicative effect-size claim — the binding signal is the absolute gap
of $0.054$ vs.\ $0.900$ on the worst cell, plus the $93\%$ paired-resample
win rate. Source: \texttt{R\_raw/cp\_baselines\_h2h/open\_targets\_*.json}.}
\label{tab:open_targets_h2h}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{lcccccc}
\toprule
Method & Marg.\ cov.\ & Cov$|Y{=}1$ & $\shat$-bin gap $\downarrow$
       & Paired Win\,\% & $p_{\mathrm{one}}$ & Singleton \\
\midrule
\textbf{HCCP (ours)} & 0.898 & $0.891$ & $\mathbf{0.054}$ & --- & --- & 0.397 \\
RLCP                 & 0.953 & \textcolor{red}{$0.815$} & \textcolor{red}{$0.900$}
                     & $93.0\%$ & $0.075$ & 0.611 \\
Weighted CP          & 0.901 & 0.902 & \textcolor{red}{$0.900$}
                     & $93.0\%$ & $0.075$ & 0.507 \\
SC-CP                & 0.900 & \textcolor{red}{$0.873$} & \textcolor{red}{$0.900$}
                     & $93.0\%$ & $0.075$ & 0.524 \\
\bottomrule
\end{tabular}
\end{table}

HCCP attains $\shat$-bin gap $0.054$ vs.\ $0.900$ for all three contemporary
baselines (the worst-cell ceiling) — a $16.7\times$ improvement at the
recommended $K_{\mathrm{eval}} = 5$;
Tab.~\ref{tab:open_targets_keval} reports a $12.9$--$16.7\times$ range across
$K_{\mathrm{eval}} \in \{2, 5\}$ within the bias regime, of comparable
magnitude to TraitGym Complex's $13.7\times$ over the on-axis baseline
(Tab.~\ref{tab:cp_baselines}). The minority-coverage advantage is also
preserved: HCCP and weighted CP both hit the $0.85$--$0.90$ band, while
RLCP and SC-CP undercover the minority class. RLCP achieves marginal
coverage $0.953 > 1 - \alpha$ at the cost of $\mathrm{cov}_{|Y=1} = 0.815$
(below the $0.85$ floor) and a wide $61.1\%$ singleton rate that exhausts
HCCP's abstention budget.

\paragraph{$K_{\mathrm{eval}}$ reversal replicates the TraitGym Mendelian
boundary.}
We sweep $K_{\mathrm{eval}} \in \{2, 3, 5, 8, 10\}$ under matching
nested-CV partition selection (Tab.~\ref{tab:open_targets_keval}). HCCP
dominates by $12.9$--$16.7\times$ at $K_{\mathrm{eval}} \in \{2, 5\}$
(per-cell-minority $\geq 228$, deep in the bias regime),
ties saturate at $K_{\mathrm{eval}} = 8$ (per-cell-minority $\approx 142$),
and \emph{reverses} at $K_{\mathrm{eval}} = 10$ where SC-CP overtakes HCCP
($0.65$ vs.\ $0.90$, paired win rate $8.0\%$, $p_{\mathrm{one}} = 0.92$ —
HCCP loses with confidence). The $K_{\mathrm{eval}} = 3$ row is anomalously
high ($0.72$) because nested-CV selected $K_{\mathrm{HCCP}} = 8$ for that
fold: the metric re-bins at $K_{\mathrm{eval}} = 3$ and aggregates over
HCCP's $K_{\mathrm{HCCP}} = 8$ partition cells, some of which trigger the
small-cell pooled fallback. The bias-variance crossover boundary on OT
sits at per-cell-minority $\approx 100$--$120$, consistent with the
$\approx 70$ floor we report on TraitGym Mendelian
(\S\ref{sec:experiments:h2h}). The minority-floor argument of
Eq.~\eqref{eq:gap_decomp}'s variance-term scaling is therefore consistent
with the regime boundary observed across two completely independent
benchmarks; the asymptotic $K^\star \approx 22$ from Eq.~\eqref{eq:gap_decomp}
itself is well above the partition granularities that bind in finite samples
(see also \S\ref{sec:theory:t5}).

\begin{table}[h]
\centering
\small
\caption{$K_{\mathrm{eval}}$ sensitivity on Open Targets matched\_9. ``HCCP win
ratio'' is best-baseline $\shat$-bin gap divided by HCCP's; values $> 1$
favour HCCP. Per-cell minority $= \pi_{+} n / K_{\mathrm{eval}}$ with
$\pi_{+} = 0.10$, $n = 11{,}400$. The $K_{\mathrm{eval}} = 3$ row reflects
a $K_{\mathrm{HCCP}} \neq K_{\mathrm{eval}}$ partition--metric mismatch
(see text); all other rows have $K_{\mathrm{HCCP}}$ modal $\in
\{5, 8, 10\}$ aligned with $K_{\mathrm{eval}}$. Excluding the $K_{\mathrm{eval}}
= 3$ mismatch row, the win ratio is monotone non-increasing in
$K_{\mathrm{eval}}$ (equivalently, in per-cell minority count), tracking
the variance-term scaling.}
\label{tab:open_targets_keval}
\setlength{\tabcolsep}{4pt}
\begin{tabular}{ccccccc}
\toprule
$K_{\mathrm{eval}}$ & per-cell min & HCCP gap & Best baseline gap
                    & HCCP win ratio & Paired win\,\% & $p_{\mathrm{one}}$ \\
\midrule
$2$  & $570$ & $\mathbf{0.036}$ & $0.468$ (SC-CP) & $\mathbf{12.9\times}$ & $\mathbf{100.0\%}$ & $\mathbf{0.005}$ \\
$3$  & $380$ & $0.718$           & $0.900$ (3-way tie) & $1.25\times^{*}$ & --- & --- \\
$5$  & $228$ & $\mathbf{0.054}$ & $0.900$ (3-way tie) & $\mathbf{16.7\times}$ & $93.0\%$ & $0.075$ \\
$8$  & $142$ & $0.900$           & $0.900$ (4-way tie) & $1.00\times$  & --- & --- \\
$10$ & $114$ & $0.900$           & \textbf{$0.650$ (SC-CP)} & $\mathbf{0.72\times}$ & $\mathbf{8.0\%}$ & $\mathbf{0.92}$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Cross-platform replication: three-axis takeaway.}
Open Targets is a feature-pipeline-disjoint, dataset-disjoint, and
benchmark-disjoint replication of the TraitGym H2H. The headline HCCP win
ratio ($16.7\times$ at $K_{\mathrm{eval}} = 5$) sits between TraitGym
Complex ($13.4\times$) and TraitGym Mendelian's contested operating point;
the $K_{\mathrm{eval}} = 10$ reversal mirrors the TraitGym Mendelian
$K_{\mathrm{eval}} \geq 5$ reversal under the same per-cell minority
threshold. We read this as direct empirical support for the
finite-sample-floor reading of Eq.~\eqref{eq:gap_decomp}: the operative
boundary between bias-dominated (HCCP wins) and variance-dominated
(weighted CP / SC-CP wins) regimes is the per-cell quantile-estimator
floor around $n_{\min} \approx 100$, not the asymptotic $\rho = 1$
crossover at $K^\star \approx \sqrt{L_F R \pi_{\min} n}$ (which is
$\geq 22$ at TraitGym / OT scales, well above any partition we operate at
in finite samples). End-to-end pipeline source:
\texttt{scripts/40\_open\_targets\_download.py} through
\texttt{scripts/44\_open\_targets\_matched\_controls.py},
\texttt{T\_tools/open\_targets\_subsample.py},
\texttt{T\_tools/cp\_baselines\_h2h.py --dataset open\_targets}.
